---
title: "Final Project"
author: "Yueyi Xu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(tidyverse)
library(glmnet)
library(caret)
library(tidymodels)
library(pls)
library(corrplot)
library(knitr)
library(patchwork)
library(tibble)
library(knitr)
library(kableExtra)
library(MASS)
library(ggplot2)
library(dplyr)
library(klaR)
library(factoextra) 
library(gridExtra) 
library(corrplot) 
library(RColorBrewer) 
library(gplots) 
library(jpeg)
library(ranger)  
library(earth)   
library(e1071)   
library(gbm)   

load("severity_test.RData")
load("severity_training.RData")
```

```{r}
training_data <- training_data %>%
  janitor::clean_names() %>%
  dplyr::select(-id) %>%
  mutate(severity = case_match(as.numeric(severity),
                               1 ~ "Not Severe",
                               2 ~ "Severe"),
         severity = factor(severity),
         gender = case_match(gender,
                             1 ~ "Male",
                             0 ~ "Female"),
         race = case_match(as.numeric(race),
                           1 ~ "White",
                           2 ~ "Asian",
                           3 ~ "Black",
                           4 ~ "Hispanic"),
         smoking = case_match(as.numeric(smoking),
                              1 ~ "Never",
                              2 ~ "Former",
                              3 ~ "Current"),
         hypertension = case_match(hypertension,
                                   0 ~ "No",
                                   1 ~ "Yes"),
         diabetes = case_match(diabetes,
                               0 ~ "No",
                               1 ~ "Yes"),
         vaccine = case_match(vaccine,
                              0 ~ "Not Vaccinated",
                              1 ~ "Vaccinated")
         )
```

```{r}
test_data <- test_data %>%
  janitor::clean_names() %>%
  dplyr::select(-id) %>%
  mutate(severity = case_match(as.numeric(severity),
                               1 ~ "Not Severe",
                               2 ~ "Severe"),
         severity = factor(severity),
         gender = case_match(gender,
                             1 ~ "Male",
                             0 ~ "Female"),
         race = case_match(as.numeric(race),
                           1 ~ "White",
                           2 ~ "Asian",
                           3 ~ "Black",
                           4 ~ "Hispanic"),
         smoking = case_match(as.numeric(smoking),
                              1 ~ "Never",
                              2 ~ "Former",
                              3 ~ "Current"),
         hypertension = case_match(hypertension,
                                   0 ~ "No",
                                   1 ~ "Yes"),
         diabetes = case_match(diabetes,
                               0 ~ "No",
                               1 ~ "Yes"),
         vaccine = case_match(vaccine,
                              0 ~ "Not Vaccinated",
                              1 ~ "Vaccinated")
         )
```



# Exploratory analysis and data visualization

We will create box plots for continuous predictors such as Age, Height, Weight, BMI, Systolic blood pressure (SBP), LDL cholesterol (LDL), and Depression. These plots will show how these metrics vary with the severity of COVID-19.

```{r}
# Boxplot for Age vs. Severity
p1 <- ggplot(training_data, aes(x = factor(severity), y = age, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "Age vs. COVID-19 Severity", x = "Severity", y = "Age") +
  scale_fill_brewer(palette = "Set1")

# Boxplot for Height vs. Severity
p2 <- ggplot(training_data, aes(x = factor(severity), y = height, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "Height vs. COVID-19 Severity", x = "Severity", y = "Height") +
  scale_fill_brewer(palette = "Set1")

# Boxplot for Weight vs. Severity
p3 <- ggplot(training_data, aes(x = factor(severity), y = weight, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "Weight vs. COVID-19 Severity", x = "Severity", y = "Weight") +
  scale_fill_brewer(palette = "Set1")

# Boxplot for BMI vs. Severity
p4 <- ggplot(training_data, aes(x = factor(severity), y = bmi, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "BMI vs. COVID-19 Severity", x = "Severity", y = "BMI") +
  scale_fill_brewer(palette = "Set1")

# Boxplot for SBP vs. Severity
p5 <- ggplot(training_data, aes(x = factor(severity), y = sbp, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "Systolic Blood Pressure vs. COVID-19 Severity", x = "Severity", y = "Systolic BP") +
  scale_fill_brewer(palette = "Set1")

# Boxplot for LDL vs. Severity
p6 <- ggplot(training_data, aes(x = factor(severity), y = ldl, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "LDL Cholesterol vs. COVID-19 Severity", x = "Severity", y = "LDL Cholesterol") +
  scale_fill_brewer(palette = "Set1")

# Boxplot for Depression vs. Severity
p7 <- ggplot(training_data, aes(x = factor(severity), y = depression, fill = factor(severity))) +
  geom_boxplot() +
  labs(title = "Depression vs. COVID-19 Severity", x = "Severity", y = "Depression") +
  scale_fill_brewer(palette = "Set1")

```

We will visualize the relationship between categorical predictors and severity. We will focus on gender, smoking status, hypertension, diabetes, and vaccination status. We will use bar plots showing the proportion within each severity category.


```{r}
# Function to create proportion bar plots
create_prop_plot <- function(data, varname) {
  plot <- data %>%
    group_by(severity, !!rlang::sym(varname)) %>%
    summarise(Count = n(), .groups = 'drop') %>%
    mutate(Prop = Count / sum(Count)) %>%
    ggplot(aes(x = factor(severity), y = Prop, fill = factor(!!rlang::sym(varname)))) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = paste(varname, "vs. COVID-19 Severity"), x = "Severity", y = "Proportion") +
    scale_fill_brewer(palette = "Set1")
  
  return(plot)
}

# Proportion Bar Plots
p8 <- create_prop_plot(training_data, "gender")
p9 <- create_prop_plot(training_data, "smoking")
p10 <- create_prop_plot(training_data, "hypertension")
p11 <- create_prop_plot(training_data, "diabetes")
p12 <- create_prop_plot(training_data, "vaccine")
```

```{r}
plots_list <- list(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12)
combined_plot <- grid.arrange(grobs = plots_list, ncol = 4, nrow = 3)

png("combined_plots.png", width = 2000, height = 1500)
grid.arrange(grobs = plots_list, ncol = 4, nrow = 3)
dev.off()
```



```{r}
dat_continuous <- training_data %>% 
  dplyr::select("age", "height", "weight", "bmi", "sbp", "ldl", "depression")
png("continuous_corrplot.png", width = 400, height = 300)

corrplot(cor(dat_continuous), method = 'number', type = 'lower') 

dev.off()
```



# Model Training

### MARS

```{r}
set.seed(1)
mars_grid <- expand.grid(degree = 1:3, 
                         nprune = 2:20)

ctrl <- trainControl(method = "cv", 
                     number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

mars.fit <- train(make.names(severity) ~ ., 
                  data = training_data, 
                  method = "earth",
                  tuneGrid = mars_grid, 
                  metric = "ROC",
                  trControl = ctrl)

ggplot(mars.fit)

mars.fit$bestTune
coef(mars.fit$finalModel)
```


### Penalized Logistic Regression

```{r}
set.seed(1)
glmnGrid <- expand.grid(.alpha = seq(0, 1, length = 21),
                        .lambda = exp(seq(-5, 1, length = 100)))

ctrl <- trainControl(method = "cv", 
                     number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

glmn.fit <- train(make.names(severity) ~.,
                  data = training_data, 
                  method = "glmnet",
                  tuneGrid = glmnGrid,
                  metric = "ROC",
                  trControl = ctrl)

myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))
plot(glmn.fit, par.settings = myPar, xTrans = function(x) log(x))

glmn.fit$bestTune
```


### SVM
```{r}
set.seed(1)
svmr.grid <- expand.grid(C = exp(seq(-5, 2, len = 50)),
                         sigma = exp(seq(-6, 1, len = 20)))

ctrl <- trainControl(method = "cv", 
                     number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

svmr.fit <- train(make.names(severity) ~ . , 
                  data = training_data,
                  method = "svmRadialSigma",
                  tuneGrid = svmr.grid,
                  metric = "ROC",
                  trControl = ctrl)

myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol)) 
plot(svmr.fit, highlight = TRUE, par.settings = myPar)
```


### Random Forest
```{r}
set.seed(1)
rf.grid <- expand.grid(mtry = 1:13,
                       splitrule = "gini",
                       min.node.size = 1:6)

ctrl <- trainControl(method = "cv", 
                     number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

rf.fit <- train(make.names(severity) ~ ., 
                data = training_data,
                method = "ranger",
                tuneGrid = rf.grid,
                metric = "ROC",
                trControl = ctrl)
rf.fit$bestTune

ggplot(rf.fit, highlight = TRUE)
```


### LDA

```{r}
set.seed(1)

ctrl <- trainControl(method = "cv", 
                     number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

lda.fit <- train(make.names(severity) ~.,
                 data = training_data, 
                 method = "lda",
                 metric = "ROC",
                 trControl = ctrl)

lda.fit$finalModel
```


### AdaBoost

```{r}
set.seed(1)
gbmA.grid <- expand.grid(n.trees = c(2000,3000,4000,5000),
                         interaction.depth = 1:3,
                         shrinkage = c(0.001,0.002,0.003),
                         n.minobsinnode = 1)

ctrl <- trainControl(method = "cv", 
                     number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

gbmA.fit <- train(make.names(severity) ~ .,
                  data = training_data,
                  method = "gbm",
                  tuneGrid = gbmA.grid,
                  metric = "ROC",
                  trControl = ctrl,
                  distribution = "adaboost",
                  verbose = FALSE)
gbmA.fit$bestTune

ggplot(gbmA.fit, highlight = TRUE)
```


### Compare all models

```{r}
resamp <- resamples(list(MARS = mars.fit,
                         GLMN = glmn.fit,
                         SVM = svmr.fit,
                         RF = rf.fit,
                         LDA = lda.fit,
                         Boosting = gbmA.fit))
summary(resamp)

bwplot(resamp, metric = "ROC")
```

```{r}
resamp_summary <- summary(resamp)

roc_summary <- as.data.frame(resamp_summary$statistics$ROC)
roc_summary$Model <- rownames(roc_summary)
write.csv(roc_summary, "Roc_summary.csv", row.names = FALSE)
```

```{r}
png("Roc_boxplot.png", width = 400, height = 300)
bwplot(resamp, metric = "ROC")
dev.off()
```

```{r}
boosting.pred <- predict(gbmA.fit, newdata = test_data, type = "prob")
boosting.pred
```



